---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

The goal of crandep is to provide functions for analysing the dependencies of CRAN packages using social network analysis.

## Installation

You can install crandep from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("clement-lee/crandep")
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(crandep)
library(dplyr)
library(ggplot2)
library(igraph)
```





## One kind of dependencies 
To obtain the information about various kinds of dependencies of a package, we can use the function `get_dep_all()` which takes the package name and the type of dependencies as the first and second arguments, respectively. Currently, the second argument accepts `Depends`, `Imports`, `LinkingTo`, `Suggests`, `Reverse_depends`, `Reverse_imports`, `Reverse_linking_to`, and `Reverse_suggests`, or any variations in their letter cases.

```{r}
get_dep_all("dplyr", "Imports")
get_dep_all("MASS", "depends")
get_dep_all("MASS", "dePends")
```

`Imports` and `Depends` are the most common types of dependencies in `R` packages, but there are other types such as `Suggests`. For more information on different types of dependencies, see [the official guidelines](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Package-Dependencies) and [http://r-pkgs.had.co.nz/description.html](http://r-pkgs.had.co.nz/description.html).





## Multiple kind of dependencies
As the information all dependencies of one package are on the same page on CRAN, to avoid scraping the same multiple times, we can use `get_dep_df()` instead of  `get_dep_all()`. The output will be a data frame instead of a character vector.

```{r}
get_dep_df("dplyr", c("imports", "LinkingTo"))
```
The column `type` is the type of the dependency converted to lower case. Also, `LinkingTo` is now converted to `linking_to` for consistency. For the four reverse dependencies, the substring `"reverse_"` will not be shown in `type`; instead the `reverse` column will be `TRUE`. This can be illustrated by the following:

```{r}
get_dep_all("abc", "depends")
get_dep_all("abc", "reverse_depends")
get_dep_df("abc", c("depends", "reverse_depends"))
```
Theoretically, for each forward dependency
```{r, echo=FALSE}
data.frame(from = "A", to = "B", type = "c", reverse = FALSE)
```
there should be an equivalent reverse dependency
```{r, echo=FALSE}
data.frame(from = "B", to = "A", type = "c", reverse = TRUE)
```
Aligning the `type` in the forward dependency and the reverse dependency enables this to be checked easily.





## Building and visualising a dependency network
To build a dependency network, we have to obtain the dependencies for multiple packages. For illustration, we choose the [core packages of the tidyverse](https://www.tidyverse.org/packages/), and find out what each package `Imports`. We put all the dependencies into one data frame, in which the package in the `from` column imports the package in the `to` column. This is essentially the edge list of the dependency network.

```{r}
df0.imports <- rbind(
    get_dep_df("ggplot2", "Imports"),
    get_dep_df("dplyr", "Imports"),
    get_dep_df("tidyr", "Imports"),
    get_dep_df("readr", "Imports"),
    get_dep_df("purrr", "Imports"),
    get_dep_df("tibble", "Imports"),
    get_dep_df("stringr", "Imports"),
    get_dep_df("forcats", "Imports")
)
head(df0.imports)
tail(df0.imports)
```

