---
title: "Modelling the number of reverse dependencies"
date: "2020-08-10"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modelling the number of reverse dependencies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

From [the dependency network of all CRAN packages](cran.html), we have seen that reverse dependencies of all CRAN packages seem to follow the power law. This echoes the phenomenon in other software dependencies observed by, for example,
[LaBelle and Wallingford, 2004](https://arxiv.org/abs/cs/0411096),
[Baxter et al., 2006](https://doi.org/10.1145/1167473.1167507),
[Jenkins and Kirk, 2007](https://doi.org/10.1016/j.ins.2007.01.021),
[Wu et al. 2007](https://ieeexplore.ieee.org/document/4362637),
[Louridas et al., 2008](https://doi.org/10.1145/1391984.1391986),
[Zheng et al., 2008](https://doi.org/10.1016/j.physa.2008.06.050),
[Kohring, 2009](https://doi.org/10.1142/S0219525909002362),
[Li et al., 2013](https://doi.org/10.1016/j.physa.2013.01.035),
[Bavota et al. 2015](https://doi.org/10.1007/s10664-014-9325-9),
and [Cox et al., 2015](https://ieeexplore.ieee.org/document/7202955).
In this vignette, we will fit the discrete power law to model this number of reverse dependencies, using functions with the suffix `upp`, namely `dupp()`, `Supp()` and `mcmc_upp()`. While we shall focus on "Imports", which is one of the serveral kinds of dependencies in R, the same analysis can be carried out for all other types.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(crandep)
library(igraph)
library(dplyr)
library(ggplot2)
```





## Obtaining the reverse dependencies
In [the vignette on all CRAN pacakges](cran.html), we looked at the dependency type `Depends`. Here, we look at the dependency type `Imports`, using the same workflow. As we are using the forward dependencies to construct the network, the number of reverse dependencies is equivalent to the in-degree of a package, which can be obtained via the function `igraph::degree()`. We then construct a data frame to hold this in-degree information.

```{r}
g0.imports <- get_graph_all_packages(type = "imports")
d0.imports <- g0.imports %>% igraph::degree(mode = "in")
df0.imports <-
    data.frame(name = names(d0.imports), degree = as.integer(d0.imports)) %>%
    dplyr::arrange(dplyr::desc(degree), name)
head(df0.imports, 10)
```

For the purpose of verification, we use the reverse dependencies to construct the network, then look at the out-degrees of the packages this time.

```{r}
g0.rev_imports <- get_graph_all_packages(type = "reverse imports")
d0.rev_imports <- g0.rev_imports %>% igraph::degree(mode = "out") # note the difference to above
df0.rev_imports <-
    data.frame(name = names(d0.rev_imports), degree = as.integer(d0.rev_imports)) %>%
    dplyr::arrange(dplyr::desc(degree), name)
head(df0.rev_imports, 10)
```

Theoretically, the two data frames are identical. Any possible (but small) difference is due to the CRAN pages being updated while scraping.

```{r}
identical(df0.imports, df0.rev_imports)
setdiff(df0.imports, df0.rev_imports)
setdiff(df0.rev_imports, df0.imports)
```





## Exploratory analysis and selecting the threshold
We construct a data frame for the empirical frequencies and survival function at the whole range of data.

```{r}
df1.imports <- df0.imports %>%
	dplyr::filter(degree > 0L) %>% # to prevent warning when plotting on log-log scale
    dplyr::count(degree, name = "frequency") %>%
	dplyr::arrange(dplyr::desc(degree)) %>%
	dplyr::mutate(survival = cumsum(frequency)/sum(frequency))
```

Before fitting the discrete power law, to be determined first is a threshold above which it is appropriate. We will visualise the degree distribution to determine such threshold.

```{r}
gg0 <- df1.imports %>%
	ggplot2::ggplot() +
	ggplot2::geom_point(aes(degree, frequency), size = 0.75) +
	ggplot2::scale_x_log10() +
	ggplot2::scale_y_log10() +
	ggplot2::coord_cartesian(ylim = c(1L, 1e+3L)) +
	ggplot2::theme_bw(12)
gg0
```

The power law seems appropriate for the whole range of data, and so, for illustration purposes, the threshold will be set at 1 inclusive. 0's will be excluded anyway because, as we will see, the probability mass function (PMF) is not well-defined at 0.

While it is straightforward to determine the threshold here, such linearity over the whole range might not be seen for other data. The package [poweRlaw](https://cran.r-project.org/package=poweRlaw) provides functions for and references to more systematic/objective procedures of selecting the threshold.





## Fitting the discrete power law
We use the function `mcmc_upp()` to fit the discrete power law, of which the PMF is proportional to $x^{-\alpha}$, where $\alpha$ is the lone scalar parameter. Here we will use the parameter $\xi_1=1/(\alpha-1)$ to align with the parameterisation of `mcmc_mix()` and other distributions in extreme value theory, which is an extension of the power law.

The Bayesian approach is used here for inference, meaning that a prior has to be set for the parameters. We assume a uniform distribution $U(a_{\xi_1}=0, b_{\xi_1}=100)$ for $\xi_1$. Markov chain Monte Carlo (MCMC) is used as the inference algorithm.

```{r,results = FALSE}
x <- dplyr::filter(df0.imports, degree > 0L)$degree # data
u <- 1L # threshold
xi1 <- 1.0 # initial value
a_xi1 <- 0.0 # lower bound of uniform distribution
b_xi1 <- 100.0 # upper bound of uniform distribution
set.seed(3075L)
mcmc0.imports <- mcmc_upp(x = x, u = u, xi1 = xi1, a_xi1 = a_xi1, b_xi1 = b_xi1) # takes seconds
```

Now we have the samples representing the posterior distribution of $\xi_1$:

```{r}
mcmc0.imports %>%
    ggplot2::ggplot() +
	ggplot2::geom_density(aes(xi1)) +
	ggplot2::theme_bw(12)
```

We can obtain the resulting posterior of $\alpha=1/\xi_1+1$.

```{r}
mcmc0.imports <- mcmc0.imports %>%
    dplyr::mutate(alpha = 1.0 / xi1 + 1.0)
mcmc0.imports %>%
	ggplot2::ggplot() +
	ggplot2::geom_density(aes(alpha)) +
	ggplot2::theme_bw(12)
```

This means the number of reverse "Imports" follows approximately a power law with exponent `r round(mean(mcmc0.imports$alpha),2)`. We can also calculate the fitted frequencies and survival function, using `dupp()` and `Supp()` respectively.

```{r}
n0 <- sum(df1.imports$frequency) # TOTAL number of data points in x
## or n0 <- length(x)
n1 <- length(df1.imports$frequency) # number of UNIQUE data points
N <- length(mcmc0.imports$xi1)
freq0 <- surv0 <- matrix(as.numeric(NA), N, n1)
for (i in seq(N)) {
    freq0[i,] <- n0 * dupp(x = df1.imports$degree, u = 1L, xi1 = mcmc0.imports$xi1[i])
    surv0[i,] <- Supp(x = df1.imports$degree, u = 1L, xi1 = mcmc0.imports$xi1[i])
}
df1.imports <- df1.imports %>%
    dplyr::mutate(
	    frequency.mean = apply(freq0, 2, mean),
		frequency.qlow = apply(freq0, 2, quantile, p = 0.025),
		frequency.qupp = apply(freq0, 2, quantile, p = 0.975),
		survival.mean = apply(surv0, 2, mean),
		survival.qlow = apply(surv0, 2, quantile, p = 0.025),
		survival.qupp = apply(surv0, 2, quantile, p = 0.975)
	)
```

Finally, we overlay the fitted line (blue, dashed) and credible intervals (red, dotted) on the plot above to check goodness-of-fit:

```{r}
gg1 <- df1.imports %>%
    ggplot2::ggplot() +
	ggplot2::geom_point(aes(degree, frequency), size = 0.75) +
	ggplot2::geom_line(aes(degree, frequency.mean), col = 4, lty = 2) +
	ggplot2::geom_line(aes(degree, frequency.qlow), col = 2, lty = 3) +
	ggplot2::geom_line(aes(degree, frequency.qupp), col = 2, lty = 3) +
	ggplot2::scale_x_log10() +
	ggplot2::scale_y_log10() +
	ggplot2::coord_cartesian(ylim = c(1L, 1e+3L)) +
	ggplot2::theme_bw(12)
gg1
```

The corresponding plot using the survival function can also be obtained:

```{r}
gg2 <- df1.imports %>%
    ggplot2::ggplot() +
	ggplot2::geom_point(aes(degree, survival), size = 0.75) +
	ggplot2::geom_line(aes(degree, survival.mean), col = 4, lty = 2) +
	ggplot2::geom_line(aes(degree, survival.qlow), col = 2, lty = 3) +
	ggplot2::geom_line(aes(degree, survival.qupp), col = 2, lty = 3) +
	ggplot2::scale_x_log10() +
	ggplot2::scale_y_log10() +
	ggplot2::theme_bw(12)
gg2
```

This shows the discrete power law doesn't fit as good as it seems according to the frequency plot. Potential improvements include using a higher threshold, which inevitably means throwing away some data, or using a more flexible distribution, such as those in extreme value theory. Provided in this package for such purposes are `dmix()`, `Smix()` and `mcmc_mix()`, which are functions for discrete extreme value mixture distributions introduced in [Lee and Eastoe, 2020](https://arxiv.org/abs/2008.03073).