---
title: "Introduction to rackage"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to rackage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette provides an introduction to using the functions to analyse the dependencies of CRAN packages using social network analysis.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(rackage)
library(dplyr)
library(ggplot2)
library(igraph)
```





## One kind of dependencies 
To obtain the information about various kinds of dependencies of a package, we can use the function `get_dep_all()` which takes the package name and the type of dependencies as the first and second arguments, respectively. Currently, the second argument accepts `Depends`, `Imports`, `LinkingTo`, `Suggests`, `Reverse_depends`, `Reverse_imports`, `Reverse_linking_to`, and `Reverse_suggests`, or any variations in their letter cases.

```{r}
get_dep_all("dplyr", "Imports")
get_dep_all("MASS", "depends")
get_dep_all("MASS", "dePends")
```

`Imports` and `Depends` are the most common types of dependencies in `R` packages, but there are other types such as `Suggests`. For more information on different types of dependencies, see [the official guidelines](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Package-Dependencies) and [http://r-pkgs.had.co.nz/description.html](http://r-pkgs.had.co.nz/description.html).





## Multiple kind of dependencies
As the information all dependencies of one package are on the same page on CRAN, to avoid scraping the same multiple times, we can use `get_dep_df()` instead of  `get_dep_all()`. The output will be a data frame instead of a character vector.

```{r}
get_dep_df("dplyr", c("imports", "LinkingTo"))
```
The column `type` is the type of the dependency converted to lower case. Also, `LinkingTo` is now converted to `linking_to` for consistency. For the four reverse dependencies, the substring `"reverse_"` will not be shown in `type`; instead the `reverse` column will be `TRUE`. This can be illustrated by the following:

```{r}
get_dep_all("abc", "depends")
get_dep_all("abc", "reverse_depends")
get_dep_df("abc", c("depends", "reverse_depends"))
```
Theoretically, for each forward dependency
```{r, echo=FALSE}
data.frame(from = "A", to = "B", type = "c", reverse = FALSE)
```
there should be an equivalent reverse dependency
```{r, echo=FALSE}
data.frame(from = "B", to = "A", type = "c", reverse = TRUE)
```
Aligning the `type` in the forward dependency and the reverse dependency enables this to be checked easily.





## Building and visualising a dependency network
To build a dependency network, we have to obtain the dependencies for multiple packages. For illustration, we choose the [core packages of the tidyverse](https://www.tidyverse.org/packages/), and find out what each package `Imports`. We put all the dependencies into one data frame, in which the package in the `from` column imports the package in the `to` column. This is essentially the edge list of the dependency network.

```{r}
df0.imports <- rbind(
    get_dep_df("ggplot2", "Imports"),
    get_dep_df("dplyr", "Imports"),
    get_dep_df("tidyr", "Imports"),
    get_dep_df("readr", "Imports"),
    get_dep_df("purrr", "Imports"),
    get_dep_df("tibble", "Imports"),
    get_dep_df("stringr", "Imports"),
    get_dep_df("forcats", "Imports")
)
head(df0.imports)
tail(df0.imports)
```

With the help of the 'igraph' package, we can use this data frame to build a graph object that represents the dependency network.

```{r, echo=c(1,2,4), out.width="660px", out.height="660px", fig.width=12, fig.height=12}
g0.imports <- igraph::graph_from_data_frame(df0.imports)
set.seed(1457L)
par(mar = rep(0.0, 4))
plot(g0.imports, vertex.label.cex = 1.5)
```

The nature of a dependency network makes it a directed acyclic graph (DAG). We can use the 'igraph' function `is_dag()` to check.

```{r}
igraph::is_dag(g0.imports)
```

Note that this applies to `Imports` (and `Depends`) only due to their nature. This acyclic nature does not apply to a network of, for example, `Suggests`.





## Boundary and giant component
It is possible to set a boundary on the nodes. 

```{r, echo=c(1,2,3,5), out.width="660px", out.height="660px", fig.width=12, fig.height=12}
df0.nodes <- data.frame(name = c("ggplot2", "dplyr", "tidyr", "readr", "purrr", "tibble", "stringr", "forcats"), stringsAsFactors = FALSE)
g0.core <- df_to_graph(df0.imports, df0.nodes)
set.seed(259L)
par(mar = rep(0.0, 4))
plot(g0.core, vertex.label.cex = 1.5)
```





## Topological ordering of nodes
Since networks according to `Imports` or `Depends` are DAGs, we can obtain the [topological ordering](https://en.wikipedia.org/wiki/Topological_sorting) using, for example, [Kahn's (1962) sorting algorithm](https://doi.org/10.1145/368996.369025).

```{r}
topo_sort_kahn(g0.core)
```
In the topological ordering, represented by the column `id_num`, a low (high) number represents being at the front (back) of the ordering. If package A `Imports` package B i.e. there is a directed edge from A to B, then A will be topologically before B. As the package 'tibble' doesn't import any package but is imported by most other packages, it naturally goes to the back of the ordering. This ordering may not be unique for a DAG, and other admissible orderings can be obtained by setting `random=TRUE` in the function:
```{r, echo=c(1)}
set.seed(387L); topo_sort_kahn(g0.core, random = TRUE)
#set.seed(106L); topo_sort_kahn(g0.core, random = TRUE)
```
We can also apply the topological sorting to the bigger dependencies network.
```{r}
df0.topo <- topo_sort_kahn(g0.imports)
head(df0.topo)
tail(df0.topo)
```





## The dependency network of all CRAN packages
Ultimately, we can use `get_dep_df()` to obtain **all** dependencies of [**all** packages available on CRAN](https://cran.r-project.org/web/packages/available_packages_by_name.html). This package provides an example dataset `cran_dependencies` that contains all such dependencies as of 2020-05-09.
```{r}
data(cran_dependencies)
cran_dependencies
```
We can build dependency network in the same way as above. We can also obtain the dependency counts for each package and each type:
```{r}
df0.summary <- cran_dependencies %>%
    dplyr::count(from, type, reverse)
df0.summary
```
We can look at the "winner" in each of the reverse dependencies:
```{r}
df0.summary %>%
    dplyr::filter(reverse) %>%
    dplyr::group_by(type) %>%
    dplyr::top_n(1, n)
```
This is not surprising given the nature of each package. To take the summarisation one step further, we can obtain the frequencies of the dependency counts. These frequencies, once normalised, represent the distribution of the dependency counts, and can be neatly visualised on the log-log scale:
```{r, out.width="660px", out.height="660px", fig.width=9, fig.height=9}
df1.summary <- df0.summary %>%
    dplyr::count(type, reverse, n)
gg0.summary <- df1.summary %>%
    dplyr::mutate(reverse = ifelse(reverse, "reverse", "forward")) %>%
    ggplot() +
    geom_point(aes(n, nn)) +
    facet_grid(type ~ reverse) +
    scale_x_log10() +
    scale_y_log10() +
    labs(x = "Dependency counts", y = "Number of packages") +
    theme_bw(20)
gg0.summary
```
This shows the reverse dependencies, in particular `Reverse_depends` and `Reverse_imports`, follow the [power law](https://en.wikipedia.org/wiki/Power-law), which is empirically observed in various academic fields.
    





## Conclusion
Methods in social network analysis, such as community detection algorithms, can be applied to study the properties of the network. With a comprehensive list of packages, a substantial dependency network can be built and visualised. Ideally, the dependencies of all the CRAN packages can be obtained to build the ultimate dependency network and to provide a bird's-eye view of the CRAN ecosystem.